Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-26 _tex/index.tex
Built site for gh-pages

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
However, despite the method's widespread use, existing literature lacks
a clear explanation of the complexities and assumptions underpinning the
DCJ system, as well as the plausible mechanisms through which DCJ data
are generated. This study addresses these issues by representing DCJ
within the framework of causal inference. Specifically, utilizing a
structural approach to causal inference, the study develops a scientific
model to clarify the causal assumptions and mechanisms inherent in the
DCJ system. It then translates this model into a probabilistic
statistical framework to estimate statistical relationships and infer
causal connections within the system. This research provides a robust
probabilistic foundation for the statistical analysis of DCJ data,
building upon Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
researchers and analysts designing and implementing DCJ experiments.
\end{abstract}

@@

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, existing literature
lacks a clear representation of the plausible mechanisms through which
DCJ data are generated. Particularly, there is no depiction of the
complexity and the assumptions underpinning the DCJ system, nor how
different assessment factors can potentially influence the observed DCJ
outcome.

@@

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing a structural
approach to causal inference
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study develops a
scientific model to clarify the causal assumptions and mechanisms
inherent in the DCJ system. Next, using a minimal set of assumptions,
the study translates the scientific model into a probabilistic
statistical model. This model aims to produce statistical estimates to
draw inferences about plausible causal relationships within the DCJ
system.

@@

According to \citet{Pearl_et_al_2018}, counterfactuals occupy the
highest level of cognitive abstraction in the ladder of causation,
followed by intervention and association, and form the foundation of
causal inference. Counterfactuals represent scenarios \emph{contrary to
fact}, where alternative \emph{potential} outcomes resulting from a
cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. Nevertheless, despite their
abstract nature, counterfactuals enable the development of a
\emph{theory of the world} that explains why specific causes have
specific effects and what occurs in their absence
\citep{Pearl_et_al_2018}. They achieve this by translating causal
statements into counterfactual statements, that is, statements about
``what would have happened in the world under different circumstances.''

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}. Both approaches
employ rigorous mathematical notation to characterize causal inference,
but they do so in different ways \citep{Neal_2020}. The potential
outcomes approach relies on counterfactual notation, whereas the
structural approach utilizes the do-operator and structural causal
models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
differences, both notations can be expressed in terms of the other, and
both approaches provide methods for using experimental and observational
data to estimate causal effects \citep{Pearl_2010}.

However, the structural approach offers a key advantage over the
potential outcomes approach: it enables the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs are heuristics that
effectively convey the assumed causal structure of a system. They do not
represent detailed statistical models but allow researchers to deduce
which statistical models can provide valid causal inferences, assuming
the causal structure depicted in the DAG is accurate
\citep{McElreath_2020}.

@@

Graph theory is the branch of mathematics focused on the study of
graphs. Graphs are mathematical structures used to model pairwise
relations between objects \citep{Gross_et_al_2018}. While graph theory
covers a wide array of topics, the field of causal inference,
particularly its structural approach, has incorporated some of its
concepts to represent causes and counterfactuals formally and
transparently. A causal graph, or Directed Acyclic Graph (DAG), as its
name suggest, is a directed graph without cycles. A \emph{graph} is a
collection of nodes connected by edges. In a \emph{directed graph},
edges extend from a node to another node, with arrows indicating the
direction of causal influence. In a directed \emph{acyclic} graph, the
direction of causal influences does not loop back on itself, ensuring
that the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.

@@

\newcommand{\dsep}{\perp\!\!\!\perp}
\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}

\begin{equation}\phantomsection\label{eq-StructuralModel}{
\begin{aligned}
  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
  \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\
  \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\
  \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\
  \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\
  & \; e_{k} \perp\!\!\!\perp e_{i} \\
  & \; e_{k} \perp\!\!\!\perp e_{r_{i}} \\
  & \; e_{i} \perp\!\!\!\perp e_{r_{i}}
\end{aligned}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-StatModel_general}{
\begin{aligned}
  O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\
  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
  \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\
  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
\end{aligned}
}\end{equation}

for identification purposes we can set
\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02\),
\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1\), and
\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1\). A special case of this
would be to assume that the data comes from the same population, in that
case, \(\sigma_{k(g)} = \sigma_{k} = 0.02\),
\(\sigma_{i(g)} = \sigma_{i} = 1\)

@@

Thurstone's comparative judgment \citet{Thurstone_1927} is based on the
formula: \[
X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}}
\]

where \(X_{AB}\) defines the comparative judgment outcome, \(S_{A}\) and
\(S_{B}\) are the modal discriminal processes,
\(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}\),
with \(\sigma_{A}\) and \(\sigma_{B}\) being the dispersion of
discriminal processes \(A\) and \(B\), respectively, and \(\rho\) the
correlation between discriminal processes.

The theory identifies five cases:

\begin{itemize}
\tightlist
\item
  \textbf{Case 1:} only constant \(\rho\) (not \(\rho_{ij}\))
\item
  \textbf{Case 2:} \(X_{ij}\) becomes \(X_{kij}\) with \(k=1, \dots, K\)
  judges (replication, not duplication)
\item
  \textbf{Case 3:} \(\rho = 0\), then
  \(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}\)
\item
  \textbf{Case 4:} \(\sigma_{B}=\sigma_{A}+d\), then
  \(\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}\)
\item
  \textbf{Case 5:} \(\sigma_{B}=\sigma_{A}\), then
  \(\sigma_{AB} = \sqrt{2}\sigma\)
\end{itemize}

Now using the DAG and statistical notation

\begin{equation}\phantomsection\label{eq-StatModel_thurstone}{
\begin{aligned}
  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
  \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\
  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
\end{aligned}
}\end{equation}

The theory identifies five cases:

\begin{itemize}
\tightlist
\item
  \textbf{Case 1:} only constant \(\rho \approx \sigma_{i}\)
\item
  \textbf{Case 2:} now judges are separated by using \(\gamma_{k}\)
\item
  \textbf{Case 3:} \(\rho \approx \sigma_{e_{i}} = 0\) (no nesting of
  texts on students), then
  \(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}\)
\item
  \textbf{Case 4:} \(\sigma_{B}=\sigma_{A}+d\), then
  \(\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}\)
\item
  \textbf{Case 5:} \(\sigma_{B}=\sigma_{A}\), then
  \(\sigma_{AB} = \sqrt{2}\sigma\)
\end{itemize}

But now can we see other scenarios than just those 5 cases?

\begin{itemize}
\tightlist
\item
  consider different \(\rho \approx \sum_{p=1}^{P} \sigma_{p}\),
  depending on \(P\) nesting structures
\item
  we can now investigate \(\gamma_{k}\)
\item
  we can assume \(\sigma_{B} \neq \sigma_{A}\), no need for results on
  the limit
\end{itemize}

>>>>

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
However, despite the method's widespread use, existing literature lacks
a clear explanation of the complexities and assumptions underpinning the
DCJ system, as well as the plausible mechanisms through which DCJ data
could be generated. This study addresses these issues by representing
DCJ within the framework of causal inference. Specifically, utilizing
the structural approach, the study develops a scientific model to
clarify the causal assumptions and mechanisms inherent in the DCJ
system. It then translates this model into a probabilistic statistical
framework to estimate statistical relationships and infer causal effects
within the system. This research provides a robust probabilistic
foundation for the statistical analysis of DCJ data, building upon
Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
Its findings offer valuable insights for researchers and analysts
designing and implementing DCJ experiments.
\end{abstract}

@@

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, existing literature
lacks a clear representation of the plausible mechanisms through which
DCJ data could be generated. Particularly, there is no depiction of the
complexity and the assumptions underpinning the DCJ system, nor how
different assessment factors can potentially influence the observed DCJ
outcome.

@@

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
develops a scientific model to clarify the causal assumptions and
mechanisms inherent in the DCJ system. Next, using a minimal set of
assumptions, the study translates the scientific model into a
probabilistic statistical model. This model aims to produce statistical
estimates to draw inferences about plausible causal relationships within
the DCJ system.

@@

Counterfactuals represent scenarios \emph{contrary to fact}, where
alternative \emph{potential} outcomes resulting from a cause are neither
observed nor observable \citep{Neal_2020, Counterfactual_2024}.
According to \citet{Pearl_et_al_2018}, counterfactuals form the
foundation of causal inference and occupy the highest level of cognitive
abstraction in the ladder of causation, followed by intervention and
association. Despite this abstract nature, counterfactuals enable the
development of a \emph{theory of the world} that explains why specific
causes have specific effects and what occurs in their absence
\citep{Pearl_et_al_2018}. They achieve this by translating causal
statements into counterfactual statements, that is, statements about
``what would have happened in the world under different circumstances.''

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}. Both approaches
employ rigorous mathematical notation to characterize causal inference,
but they do so in different ways \citep{Neal_2020}. The potential
outcomes approach relies on counterfactual notation, whereas the
structural approach employs the do-operator and structural causal models
\citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these differences,
both notations can be expressed in terms of the other, and both
approaches provide methods for using experimental and observational data
to estimate causal effects \citep{Pearl_2010}.

Nevertheless, the structural approach offers an additional key advantage
over the potential outcomes approach: it enables the graphical
representation of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs are heuristics that
can effectively convey the assumed causal structure of a system. They do
not represent detailed statistical models but allow researchers to
deduce which statistical models can provide valid causal inferences,
assuming the causal structure depicted in the DAG is accurate
\citep{McElreath_2020}.

@@

@@

@@

@@
